\section{Validity}
\label{sec:validity}

The validity of the Effect Propagation Process must be assessed according to the nature of its contribution. The EPP is
a philosophy-informed, formal computational framework designed to provide a new, more expressive language for modeling
dynamic, contextual causal systems. Therefore, its validity rests on the criteria appropriate for such a framework,
analogous to how one would assess a mathematical logic or a new programming language:

\begin{itemize}
    \item \textbf{Internal Validity} (Soundness \& Consistency): The framework's internal validity is determined by its logical and  mathematical soundness and coherence. As detailed in the accompanying specification, the EPP is built on a foundation  of first-principles reasoning to ensure its components are consistent and its operations are sound.
  \item \textbf{External Validity} (Expressive Power \& Scoped Generalization): The framework's external validity is demonstrated by its
  robustness against alternative interpretations of its premises and how well it defines the boundaries of its
  generalization.
\end{itemize}


\subsection{Internal validity}
\label{sec:validity_internal}

The \textbf{soundness} of EPP derives from the first-principle logical progression of the presented argument. The stated problem of inadequacy of classical causality in light of the challenges imposed by Quantum Gravity (emergent spacetime, indefinite causal order) is well-recognized and documented\cite{MriniHardyIndefinite}.
From this recognized problem, the argument for EPP progresses as stated below:

\begin{enumerate}
    \item Establishes classical causality and its historical critiques.
    \item Introduces the challenges from modern physics (GR, QG).
    \item Shows why these challenges render classical causality insufficient.
    \item Proposes EPP as a coherent response to these challenges.
    \item Contrasts EPP with classical causality.
    \item Discusses its ontological, epistemological, and teleological implications.
    \item Acknowledges and addresses threats to its validity.
\end{enumerate}

The soundness is further strengthened by its inspiration from established scientific theories (QFT, GR) and foundational work in quantum gravity. While quantum gravity as a scientific theory remains a work in progress, the conceptual challenges that arise from it are valid regardless of how any particular theory may explain the underlying quantum mechanisms.

The \textbf{consistency} of the EPP framework arises from a handful of carefully stated conclusions.

\subsubsection{Spacetime Agnosticism \& Causaloids}

\textbf{Premise:} On a quantum level, spacetime may not exist.

\textbf{Conclusion:} Therefore, remove spacetime from EPP.

\textbf{Premise:} EPP does not have a defined spacetime.

\textbf{Conclusion:} Cause and effect cannot be separated anymore because there is no a priori temporal order.

\textbf{Premise:} Cause and effect cannot be discerned because of missing temporal order.

\textbf{Conclusion:} Fold cause and effect into one entity, the causaloid, that is independent of temporal (and spatial) order.

Note, the last conclusion holds because of the temporal order required for classical causality. The only logical alternative conclusion from the premise of missing temporal order would be to abandon causality altogether. However, this conflicts with the reality in which causal relationships indeed exist; therefore, the alternative has been deemed unsound.

\subsubsection{Effect Propagation as the Essence of Causality}

\textbf{Premise:} The causaloid, as a building block, is independent of temporal (and spatial) order.

\textbf{Conclusion:} Define causality by what it does (propagate effects) instead of what it was thought to be (a temporal-order-dependent atomic relationship).

The careful reader may raise a concern over the choice of words (i.e., propagate effects), since “transferring information” or “transmitting influence” might be equally valid choices. True, that is indeed a fair point. However, the term information has specific meaning in information theory and computer science, and likewise, the term influence has specific meaning in social science; therefore, the author settled cautiously on “effect” mainly to prevent conflating different meanings.

\subsubsection{Compatibility with Classical Causality}

The full logical argument of how classical causality is derived from EPP is in the section “Causality as Effect Propagation Process”.
While classical causality can be derived from EPP, the reverse is not true because EPP cannot be derived from classical causality because classical causality requires a background spacetime. That deduction proves that EPP is more abstract in the sense of more general than classical causality. Therefore, it follows that EPP naturally applies to areas where classical causality cannot be used any longer. The internal validity of EPP roots in its internal soundness and consistency that stems from its first-principles reasoning. Therefore, ambiguity, contradictory claims, and unjustified leaps in logic are avoided to the extent it is possible. Minor mistakes might be possible and the author is open to suggestions for improving the EPP further.

\subsection{External validity}
\label{sec:validity_external}

Establishing the boundaries of generalization as a proxy for external validity requires a delicate balance of realistically acknowledging what EPP can address versus avoiding overstating  any particular capability. Related to external validity is always the possibility of an alternative interpretation of the underlying premises.


\subsubsection{Falsifiability}

A crucial distinction must be made regarding the role of falsifiability. The EPP, as a formal foundation, is not itself an empirical scientific theory and therefore is not directly falsifiable. One cannot falsify the established philosophical foundation of the EPP—its metaphysics, ontology, and epistemology—without challenging its underlying axiomatic assumptions. Instead, the validity of the EPP rests on its internal consistency and its expressive power.

Falsifiability, however, is a critical property of the \textbf{specific, testable causal models that are constructed \textit{within} the EPP.} A model of an avionics system, for instance, represents a set of falsifiable hypotheses encoded as a `CausaloidGraph` and its associated `Context`. This model makes concrete predictions about system behavior that can be tested against simulation or real-world data. If the predictions fail, the \textit{model} is falsified, not the framework.

The EPP provides the formal language and the computational primitives necessary to empower engineers and domain experts to express and test causal hypotheses in complex, contextual realities that were previously beyond the reach of formal modeling.

\subsubsection{Alternative interpretations}

There are several potential alternative interpretations of the premises underlying the Effect Propagation Process framework.

\textbf{Russell was more right than acknowledged}

One can take the position that, if Russell deemed causality as a relic of a bygone era, then why not openly ask to abandon causality altogether and focus solely on descriptive and correlation-based data science?

DARPA disagrees\cite{DARPA_ANSR}:

\begin{quote}
    “In the real world, observations are often correlated and a product of an underlying causal mechanism, which can be modeled and understood.”
\end{quote}

The problem is not a simple choice between correlation and causation. The deeper issue, as contemporary philosophers
like Luciano Floridi have argued, is that the predominant mental model underlying complex system design remains rooted
in an outdated Aristotelian and Newtonian "Ur-philosophy" of fixed space and time. This leads to tools that are
inadequate for the intricate complexity of reality. To address this core critique, the EPP lifts causality into a
contextual generalization required to model dynamic systems that no longer adhere to a fixed background spacetime. This
includes systems with non-Euclidean geometries, multiple interacting contexts, and emergent causal structures, as found
in domains like avionics. The question of falsifiability, therefore, applies not to the EPP framework itself, but to the
specific, testable models that are constructed within it. A model of an avionics system built using the EPP is indeed
falsifiable against simulation data; the framework itself simply provides the language for expressing that model.

The author argues that a similar shift towards a richer contextualization of advanced dynamic models needs to happen in the
correlation-based methodologies of deep learning as well, to build tools more suitable for today's reality. A
transfer of core EPP concepts, i.e., the contextual hypergraph, uniform Euclidean and non-Euclidean geometries, into the
foundations of deep learning is welcomed by the author.

\textbf{Pluralism of causal concepts}

Instead of a unified framework like EPP, one can argue in favor of the existing pluralistic reality where multiple 
disjoint concepts of (computational) causality exist for different levels of causal analysis.

As stated before, EPP does not seek to replace classical causality and all tools that are built atop the classical
definition of causality. Instead, EPP seeks to advance the core concept of causality to meet increasingly challenging
demands. Due to the novelty of this foundational work, a single coherent framework is preferred until it becomes clear
which parts may branch out and become more specialized domains over time.


\subsubsection{Boundaries of Generalization}

The primary boundary of the Effect Propagation Process is one of scope: it is a foundational framework, not a final, comprehensive theory of all dynamic causal phenomena. 
Its purpose is to provide the formal language and computational primitives necessary to begin a systematic exploration of dynamic causality, not to claim to have concluded it. 
The EPP foundation defines two clear boundaries:


\begin{itemize}
    \item \textbf{A Lower Boundary of Complexity:} The EPP is designed for a class of problems characterized by
    nonlinear temporality, non-Euclidean structures, multiple contexts, and the potential for emergence. For systems
    that align well with the classical assumptions of a fixed spacetime and static causal rules, simpler, established
    methodologies such as Pearl's SCMs are often more appropriate and should be preferred. The EPP's power is best
    reserved for the complex realities that lie beyond the classical scope.

    \item \textbf{An Upper Boundary of Knowledge:} The EPP does not claim to have "solved" dynamic causality. On the
    contrary, by providing the first formal tools to model phenomena like causal/contextual co-emergence, it reveals
    the vastness of our current ignorance. The framework, in its early stages, makes it possible to ask new kinds of
    questions and to begin building and testing the first generation of truly dynamic causal theories for which the 
    specific characteristics need to be discovered.
\end{itemize}

While the EPP draws conceptual inspiration from physics to address these new frontiers, it is a foundational and formal computational
framework. It does not propose that macroscopic systems are quantum mechanical, but rather draws inspiration from these powerful
concepts to engineer a more robust and flexible toolkit for modeling the complex, adaptive systems we face today.

\newpage

\subsubsection{Method Selection Criteria: Classical Causality vs. EPP}

In cases where methods of classical causality and conventional machine learning do not solve the problem at hand, methodologies rooted in EPP might be preferred. The following decision matrix supports the assessment of when to use which methods. This matrix provides guidance on selecting an appropriate causal modeling approach based on the temporal and spatial complexity of the system under investigation.

% Please add the following required packages to your document preamble:
\begin{table}[hb]
\begin{tabular}{llll}
Feature Assessed &
  System Characteristic &
  Classical Methods &
  EPP Methods \\ \hline
\multicolumn{1}{|l|}{Temporal Complexity} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Single timescale + linear progression} &
  \multicolumn{1}{l|}{Sufficient} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Multiple timescales OR nonlinear temporal relationships} &
  \multicolumn{1}{l|}{May struggle} &
  \multicolumn{1}{l|}{Consider EPP} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Multiple timescales AND nonlinear temporal relationships} &
  \multicolumn{1}{l|}{Insufficient} &
  \multicolumn{1}{l|}{EPP Required} \\ \hline
\multicolumn{1}{|l|}{Spatial Structure} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Euclidean space with fixed coordinates} &
  \multicolumn{1}{l|}{Sufficient} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Non-Euclidean OR dynamic spatial relationships} &
  \multicolumn{1}{l|}{Limited/Difficult} &
  \multicolumn{1}{l|}{EPP Advantageous} \\ \hline
\multicolumn{1}{|l|}{} &
  \multicolumn{1}{l|}{Non-Euclidean AND dynamic spatial relationships} &
  \multicolumn{1}{l|}{Very Limited} &
  \multicolumn{1}{l|}{EPP Required} \\ \hline
\multicolumn{1}{|l|}{Combined Complexity} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{Scenario 1} &
  \multicolumn{1}{l|}{Linear Time \& Euclidean Space} &
  \multicolumn{1}{l|}{Preferred} &
  \multicolumn{1}{l|}{} \\ \hline
\multicolumn{1}{|l|}{Scenario 2} &
  \multicolumn{1}{l|}{Nonlinear Time OR Complex Spatial} &
  \multicolumn{1}{l|}{Limited/Difficult} &
  \multicolumn{1}{l|}{Consider EPP} \\ \hline
\multicolumn{1}{|l|}{Scenario 3} &
  \multicolumn{1}{l|}{Nonlinear Time AND Complex Spatial} &
  \multicolumn{1}{l|}{Insufficient} &
  \multicolumn{1}{l|}{EPP Required} \\ \hline
\end{tabular}
\caption{Causal Method Selection Matrix}
\label{tab:method_matrix}
\end{table}

\textbf{Emergent Causality:} If the problem involves emergent causal structures (where the causal graph itself is not fixed and changes dynamically based on context, i.e., dynamic regime shifts), EPP-based methodologies become the only viable option.


\clearpage
